Finally Tagless, Partially Evaluated

### 2008-02-19 tagless/tagless-final.tex
Removed section 4.3 with the description of GADT. That section was removed from the APLAS final paper too

Roughly, a partial evaluator interprets each object term to yield either
a static (present-stage) term (using\ifshort\else\ the evaluator\fi~|R|) or
a dynamic (future-stage) term (using\ifshort\else\ the
compiler\fi~|C|). To
distinguish between static and dynamic terms, we might try to define
|repr| in the partial evaluator as
\texttt{type ('c,'dv)\,repr 
	= S0 of ('c,'dv)\,R.repr~}\Verb+|+\texttt{~E0 of ('c,'dv)\,C.repr}.
Integer and boolean literals are immediate, present-stage
values. Addition yields a static term (using~|R.add|) if and only 
if both operands are static; otherwise we extract the dynamic terms 
from the operands and add them using~|C.add|. We use |C.int| to
convert from the static term |('c,int) R.repr|, which is just |int|,
to the dynamic term.

@@

Let us re-examine the problem in~\S\ref{S:PE-problem}. What we
would ideally like is to write
\begingroup\sloppy
\texttt{type ('c,'dv) repr = P1 of (repr\_pe ('c,'dv)) R.repr option * ('c,'dv) C.repr}
where |repr_pe| is the type function defined
% inductively because P below depends on repr_pe
by
\begin{code3}
repr_pe ('c,int) = ('c,int); repr_pe ('c,bool) = ('c,bool)
repr_pe ('c,'a->'b) = ('c,'a) repr -> ('c,'b) repr
\end{code3}
\endgroup\noindent
Although we can use type classes to define this type function
in Haskell, that is not portable to MetaOCaml. However,
these three typecase alternatives are already present in existing
methods of |Symantics|.
A simple and portable solution thus emerges: we bake |repr_pe| 
into the signature |Symantics|. 
We recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that the |repr| type
constructor took two arguments |'c| and~|'dv|. We add an argument
|'sv| for the result of applying |repr_pe| to~|'dv|.
Figure~\ref{fig:ocaml} shows the new signature.
% aplas: abbreviated figure
\begin{figure}
\begin{floatrule}
\ifshort
\smallskip
\begin{code2}
module type Symantics = sig type ('c,'sv,'dv) repr
  val int : int -> ('c,int,int) repr
  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
            -> ('c,'x,'da -> 'db) repr
  val app : ('c,'x,'da -> 'db) repr
            -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
  val fix : ('x -> 'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr,
                                'da -> 'db) repr as 'x)
  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val if_ : ('c,bool,bool) repr
            -> (unit->'x) -> (unit->'x) -> (('c,'sa,'da) repr as 'x) end
\end{code2}
\medskip
\else
\begin{code2}
module type Symantics = sig
  type ('c,'sv,'dv) repr

  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val mul : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val leq : ('c,int,int) repr -> ('c,int,int) repr -> ('c,bool,bool) repr
  val if_ : ('c,bool,bool) repr
            -> (unit -> 'x) -> (unit -> 'x) -> (('c,'sa,'da) repr as 'x)
end
\end{code2}
\fi
\end{floatrule}
\caption{A (Meta)OCaml embedding of our object language that supports
  partial evaluation \ifshort \protect\linebreak[2] (\texttt{bool},
  \texttt{mul}, \texttt{leq} are elided)\fi}
\label{fig:ocaml}
\end{figure}

\begin{figure}
\begin{floatrule}
\ifshort
\begin{code2}
module P = struct
  type ('c,'sv,'dv) repr = {st: 'sv option; dy: ('c,'dv) code}
  let abstr {dy = x} = x    let pdyn x = {st = None; dy = x}

  let int  (x:int)  = {st = Some (R.int x);  dy = C.int x}
  let add e1 e2 = match e1, e2 with
  | {st = Some 0}, e | e, {st = Some 0} -> e
  | {st = Some m}, {st = Some n} -> int (R.add m n)
  | _ -> pdyn (C.add (abstr e1) (abstr e2))
  let if_ eb et ee = match eb with
  | {st = Some b} -> if b then et () else ee ()
  | _ -> pdyn (C.if_ (abstr eb) (fun () -> abstr (et ()))
                                (fun () -> abstr (ee ())))
  let lam f = {st = Some f; dy = C.lam (fun x -> abstr (f (pdyn x)))}
  let app ef ea = match ef with {st = Some f} -> f ea
                  | _ -> pdyn (C.app (abstr ef) (abstr ea)) end
\end{code2}
\medskip
\else
\begin{code2}[commandchars=\@\[\]]
module P = struct
  type ('c,'sv,'dv) repr = {st: 'sv option; dy: ('c,'dv) code}
  let abstr {dy = x} = x
  let pdyn x = {st = None; dy = x}

  let lam f = {st = Some f; dy = C.lam (fun x -> abstr (f (pdyn x)))}
  let app ef ea = match ef with
                  | {st = Some f} -> f ea
                  | _ -> pdyn (C.app (abstr ef) (abstr ea))
  let fix f = let fdyn = C.fix (fun x -> abstr (f (pdyn x)))
              in let rec self = function
                                | {st = Some @!_} as e -> app (f (lam self)) e
                                | e -> pdyn (C.app fdyn (abstr e))
                 in {st = Some self; dy = fdyn}
end
\end{code2}
\fi
\end{floatrule}
\caption{Our partial evaluator (\ifshort \texttt{bool}, \texttt{mul},
  \texttt{leq} and \texttt{fix} \else \texttt{mul} and \texttt{leq} \fi
  are elided)}
\label{fig:pe}
\end{figure}

@@

Any partial evaluator must decide how much to unfold recursion
statically: unfolding too little can degrade the residual code, whereas
unfolding too much risks nontermination.  Our partial evaluator is no
exception, because our object language includes |fix|.  The code in
Figure~\ref{fig:pe} takes the na\"\i ve approach of ``going all the
way'', that is, whenever the 
argument is static, it unfolds |fix| rather than residualize it.
\ifshort
In the accompanying source code is a conservative
alternative |P.fix| that unfolds recursion only once, then residualizes.
\else
A conservative alternative is to unfold recursion only once, then residualize:
\begin{code}
let fix f = f (pdyn (C.fix (fun x -> abstr (f (pdyn x)))))
\end{code}
\fi
Many sophisticated approaches have been developed to decide how much to unfold
\citep{jones-partial}, but this issue is orthogonal to our presentation.
\ifshort\else
A separate concern in our treatment of |fix| is possible code bloat in
the residual program, which calls for let-insertion
\citep{SwadiTahaKiselyovPasalic2006}.
\fi

@@

\section{A tagless partial evaluator}\label{PE}
Surprisingly, we can write a partial evaluator using the idea above,
namely to build object terms using ordinary functions rather than data
constructors.  We present this partial evaluator in a sequence of four
attempts. It uses no universal type and no tags
for object types.  We then discuss residualization and binding-time
analysis.  Our partial evaluator is a modular extension of the evaluator
in~\S\ref{S:interpreter-RL} and the compiler in~\S\ref{S:compiler}, in
that it uses the former to reduce static terms and the latter to build
dynamic terms.

@@

Roughly, a partial evaluator interprets each object term to yield either
a static (present-stage) term (using the evaluator~|R|) or
a dynamic (future-stage) term (using the compiler~|C|).  To
distinguish between static and dynamic terms, one might define
\begin{code}
data P0 t = S0 (R t) | E0 (C t)
\end{code}
To extract a dynamic term from this type, we create the functions
\begin{code}
abstrI :: P0 Int -> C Int
abstrI (S0 r) = int (unR r)
abstrI (E0 c) = c

abstrB :: P0 Bool -> C Bool
abstrB (S0 r) = bool (unR r)
abstrB (E0 c) = c
\end{code}
We then start to define |P0| as an instance of the |Symantics| class.
Integer and boolean literals are immediate, present-stage
values. Addition yields a static term (using~|R|) if and only if both operands
are static; otherwise we extract the dynamic terms from the operands and
add them using~|C|.
\begin{code}
instance Symantics P0 where
  int  x = S0 (int x)
  bool x = S0 (bool x)
  add (S0 e1) (S0 e2) = S0 (add e1 e2)
  add     e1      e2  = E0 (add (abstrI e1) (abstrI e2))
\end{code}
Thus, the two uses of |add| above refer to
|add| for~|R| and |add| for~|C|.

Whereas |mul| and |leq| are as easy to define as |add|, we encounter
a problem with |if_|.  If the first argument to |if_| is a dynamic term
(of type |C Bool|), the second a static term (of type |R a|) and the third a
dynamic term (of type |C a|), then we need to convert
the static term to dynamic, but there is no polymorphic ``lift''
function, of type |a -> C a|, to send a value to the future stage
\citep{xi-guarded,WalidPOPL03}. (By the way, if we
were to add polymorphic \texttt{lift} to the type class
\texttt{Symantics repr}, then \texttt{repr} would become an instance of
\texttt{Applicative} and thus \texttt{Functor}:\texttt{ fmap
f = app (lift f)~}.)

Our |Symantics| class only includes separate lifting methods |bool| and
|int|, not a parametrically polymorphic lifting method, for good reason:
When compiling to a first-order target language such as machine code,
booleans, integers, and functions may well be represented differently.
Thus, compiling polymorphic lift requires intensional type
analysis.  To avoid needing polymorphic lift, we turn to
Asai's technique \citep{asai-binding-time,sumii-hybrid}:
build a dynamic term
alongside every static term.

@@

\begin{SaveVerbatim}{P1}
data P1 t = P1 (Maybe (R t)) (C t)
\end{SaveVerbatim}

We switch to the Haskell data type
\UseVerbatim{P1}
so that a partially evaluated term |P1 t| always contains a dynamic
component and sometimes contains a static component.  By 
distributivity, the two
alternative constructors of a |Maybe| value, |Just| and |Nothing|,
tag each partially evaluated term with a phase: either present or
future.  This tag is not an object type tag: all pattern matching below
is exhaustive.  Because the future-stage component is always available, we
can now define the polymorphic function
\begin{code}
abstr1 :: P1 t -> C t
abstr1 (P1 _ dyn) = dyn
\end{code}
to extract it without requiring polymorphic lift into~|C|.  We then try
to define the interpreter |P1|---and get as far as the first-order
constructs of our object language, including |if_|.
\begin{code}
instance Symantics P1 where
  int  x = P1 (Just (int  x)) (int  x)
  bool b = P1 (Just (bool b)) (bool b)

  add (P1 (Just n1) _) (P1 (Just n2) _) = int (unR (add n1 n2))
  add e1 e2 = P1 Nothing (add (abstr1 e1) (abstr1 e2))
  -- mul and leq are analogous and elided

  if_ (P1 (Just s) _) et ef = if unR s then et else ef
  if_ eb et ef = P1 Nothing
                    (if_ (abstr1 eb) (abstr1 et) (abstr1 ef))
\end{code}
When we come to functions, however, we stumble.  According to our
definition of~|P1|, a partially evaluated object function, such as the
identity $\fun{x}x$ embedded in Haskell as |lam (\x -> x)|\texttt{ ::
}|P1 (a -> a)|, consists of a dynamic part (type |C (a -> a)|) and
maybe a static part (type |R (a -> a)|).  The dynamic part is useful
when this function is passed to another function that is only
dynamically known, as in $\fun{k}k(\fun{x}x)$.  The static part is
useful when this function is applied to a static argument, as in
$(\fun{x}x)\True$.  Neither part, however, lets us \emph{partially}
evaluate the function, that is, compute as much as possible statically
when it is applied to a mix of static and dynamic inputs.  For example,
the partial evaluator should turn $\fun{n}(\fun{x}x)n$ into $\fun{n}n$
by substituting $n$ for~$x$ in the body of $\fun{x}x$ even though $n$ is
not statically known.  The same static function, applied to
different static arguments, can give both static and dynamic results: we
want to simplify $(\fun{y}x\times y)0$ to~$0$ but $(\fun{y}x\times y)1$
to~$x$.

To enable these simplifications, we delay binding-time analysis
for a static function until it is applied, that is, until |lam f|
appears as the argument of |app|.  To do so, we have to incorporate |f|
as it is into the |P1| data structure: applying the type constructor
|P1| to a function type |a -> b| should yield one of
\begin{code}
data P1 (a -> b) = S1 (P1 a -> P1 b) | E1 (C (a -> b))
data P1 (a -> b) = P1 (Maybe (P1 a -> P1 b)) (C (a -> b))
\end{code}
unlike applying |P1| to |Int| or |Bool|.
That is, we need a nonparametric data type, something akin to
type-indexed functions and type-indexed types, which
\citet{oliveira-typecase} dub the \emph{typecase} design pattern.
Thus, typed partial evaluation, like typed CPS transformation,
inductively defines a map from source types to target types that
performs case distinction on the source type. The connection with CPS
is not an accident, as we shall see in~\S\ref{S:CPS}.

\subsection{Eliminating tags from typecase}
\label{S:PE-GADT}

Two common ways to provide typecase in Haskell are
GADTs and type-class functional dependencies
\citep{oliveira-typecase}.  These
methods are equivalent, and here we use GADTs; |incope1.hs|
in the accompanying source code shows the latter.
We introduce a GADT with four data constructors.
\begin{code}
data P t where
  VI :: Int  -> P Int
  VB :: Bool -> P Bool
  VF :: (P a -> P b) -> P (a -> b)
  E  :: C t -> P t
\end{code}
The constructors |VI|, |VB|, and |VF| build static terms (like |S0|
in~\S\ref{S:PE-lift}), and |E| builds dynamic terms (like |E0|).  However,
the type |P t| is no longer parametric in~|t|: the constructor |VF| takes an
operand of type |P a -> P b| rather than |a -> b|.  We define a function
like |abstr1| above to extract a future-stage computation from a 
value of type |P t|.
\begin{code}
abstr :: P t -> C t
abstr (VI i) = int i
abstr (VB b) = bool b
abstr (VF f) = lam (abstr . f . E)
abstr (E x)  = x
\end{code}
The cases of this function |abstr| are type-indexed.  In particular, the |VF f|
case uses the method |lam| of the |C| interpreter to compile~|f|.

We may now make |P| an instance of
|Symantics| and implement the partial evaluator as follows. We elide
|mul|, |leq|, |if_|, and |fix|.
\begin{code}
instance Symantics P where
  int x               = VI x
  bool b              = VB b
  add (VI n1) (VI n2) = VI (n1 + n2)
  add e1 e2           = E (add (abstr e1) (abstr e2))
  lam                 = VF
  app (VF f) ea       = f ea
  app (E f)  ea       = E (app f (abstr ea))
\end{code}
The implementations of |int|, |bool|, and |add| are like 
in~\S\ref{S:PE-problem}. The interpretation of |lam f| is |VF f|, 
which just wraps the HOAS function |f|. 
We can always compile |f| to a code value,
but we delay it to apply |f| to concrete arguments. The interpretation of
|app ef ea| checks to see if |ef| is such a delayed
HOAS function |VF f|. If it is, we apply |f| to the
concrete argument |ea|, giving us a chance to perform static
computations (see example |testpowfix7| in~\S\ref{S:PE-solution}). If |ef| is a
dynamic value |E f|, we residualize.

This solution using GADTs works but is not quite satisfactory. First, it
cannot be ported to MetaOCaml, as GADTs are unavailable there.  Second,
the problem of nonexhaustive pattern\hyp matching reappears in
|app| above: the type |P t| has four constructors, of which the pattern in
|app| matches only |VF| and~|E|. One may say that the
constructors |VI| and |VB| obviously cannot occur because they do not
construct values of type |P (a -> b)| as required by the type of |app|.
Indeed, the metalanguage implementation could reason thus:
if we use inductive families (as in Coq) or logical
  frameworks with canonical forms (as in Twelf with its coverage checker),
  we can prove the pattern matching to be exhaustive.
Then again, the metalanguage implementation may not reason thus:
GHC cannot and issues warnings.
Although this point may seem minor, it is the heart of
the tagging problem and the purpose of tag elimination. A typed tagged
interpreter contains many pattern\hyp matching forms that look partial
but never fail in reality. The
goal is to make this exhaustiveness \emph{syntactically} apparent.

@@

Let us re-examine the problem in~\S\ref{S:PE-problem}. 
\oleg{If we drop the previous Section, the phrase above should be changed.}
What we
would ideally like is to write
\begin{code}
data P t = P (Maybe (repr_pe t)) (C t)
\end{code}
where |repr_pe| is the type function defined
% inductively because P below depends on repr_pe
by
\begin{code}
repr_pe Int      = Int
repr_pe Bool     = Bool
repr_pe (a -> b) = P a -> P b
\end{code}
Although we can use type classes to define this type function
in Haskell, that solution is not portable to MetaOCaml. However,
these three typecase alternatives are already present in existing
methods of |Symantics|.
A simple and portable solution thus emerges: we bake |repr_pe| 
into the signature |Symantics|. Switching to MetaOCaml to demonstrate
this portability, we recall from Figure~\ref{fig:ocaml-simple} 
in~\S\ref{encoding} that the |repr| type
constructor took two arguments |'c| and~|'dv|. We add an argument
|'sv| for the result of applying |repr_pe| to~|'dv|.
Figure~\ref{fig:ocaml} shows the new signature.
\begin{figure}
\begin{floatrule}
\begin{code2}
module type Symantics = sig
  type ('c,'sv,'dv) repr

@@

  let lam f = {st = Some f; dy = C.lam (fun x -> abstr (f (pdyn x)))}
  let app ef ea = match ef with
                  | {st = Some f} -> f ea
                  | _ -> pdyn (C.app (abstr ef) (abstr ea))
  let fix f = let fdyn = C.fix (fun x -> abstr (f (pdyn x)))
              in let rec self = function
                                | {st = Some @!_} as e -> app (f (lam self)) e
                                | e -> pdyn (C.app fdyn (abstr e))
                 in {st = Some self; dy = fdyn}
end
\end{code2}
\end{floatrule}
\caption{Our partial evaluator \texttt{mul} and \texttt{leq}
  are elided)}
\label{fig:pe}
\end{figure}

@@

Figure~\ref{fig:pe} shows the partial evaluator~|P|.
Its type |repr| literally expresses the type equation for |repr_pe| above.
The function |abstr|, as in~\S\ref{S:PE-GADT},
extracts a future-stage code value from the result of
partial evaluation.  Conversely, the function |pdyn| injects a
code value into the |repr| type. As
in~\S\ref{S:PE-problem}, we build dynamic terms alongside
any static ones to avoid polymorphic lift.

>>>>

Roughly, a partial evaluator interprets each object term to yield either
a static (present-stage) term (using\ifshort\else\ the evaluator\fi~|R|) or
a dynamic (future-stage) term (using\ifshort\else\ the
compiler\fi~|C|). To
distinguish between static and dynamic terms, we might try to define
|repr| in the partial evaluator as
\texttt{type ('c,'dv)\,repr 
	= S0 of ('c,'dv)\,R.repr~}\Verb+|+\texttt{~D0 of ('c,'dv)\,C.repr}.
Integer and boolean literals are immediate, present-stage
values. Addition yields a static term (using~|R.add|) if and only 
if both operands are static; otherwise we extract the dynamic terms 
from the operands and add them using~|C.add|. We use |C.int| to
convert from the static term |('c,int) R.repr|, which is just |int|,
to the dynamic term.

@@

Let us re-examine the problem in~\S\ref{S:PE-problem}. What we
would ideally like is to write
\begingroup\sloppy
\texttt{type ('c,'dv) repr = P1 of (repr\_pe ('c,'dv)) R.repr option * ('c,'dv) C.repr}
where |repr_pe| is the type function defined
% inductively because P below depends on repr_pe
by
\begin{code3}
repr_pe ('c,int) = ('c,int); repr_pe ('c,bool) = ('c,bool)
repr_pe ('c,'a->'b) = ('c,'a) repr -> ('c,'b) repr
\end{code3}
\endgroup\noindent
Although we can use type classes to define this type function
in Haskell, that is not portable to MetaOCaml. However,
these three typecase alternatives are already present in existing
methods of |Symantics|.
A simple and portable solution thus emerges: we bake |repr_pe| 
into the signature |Symantics|. 
We recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that the |repr| type
constructor took two arguments |'c| and~|'dv|. We add an argument
|'sv| for the result of applying |repr_pe| to~|'dv|.
Figure~\ref{fig:ocaml} shows the new signature.
\begin{figure}
\begin{floatrule}
\smallskip
\begin{code2}
module type Symantics = sig type ('c,'sv,'dv) repr
  val int : int -> ('c,int,int) repr
  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
            -> ('c,'x,'da -> 'db) repr
  val app : ('c,'x,'da -> 'db) repr
            -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
  val fix : ('x -> 'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr,
                                'da -> 'db) repr as 'x)
  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val if_ : ('c,bool,bool) repr
            -> (unit->'x) -> (unit->'x) -> (('c,'sa,'da) repr as 'x) end
\end{code2}
\medskip
\end{floatrule}
\caption{A (Meta)OCaml embedding of our object language that supports
  partial evaluation \ifshort \protect\linebreak[2] (\texttt{bool},
  \texttt{mul}, \texttt{leq} are elided)\fi}
\label{fig:ocaml}
\end{figure}

\begin{figure}
\begin{floatrule}
\begin{code2}
module P = struct
  type ('c,'sv,'dv) repr = {st: 'sv option; dy: ('c,'dv) code}
  let abstr {dy = x} = x    let pdyn x = {st = None; dy = x}

  let int  (x:int)  = {st = Some (R.int x);  dy = C.int x}
  let add e1 e2 = match e1, e2 with
  | {st = Some 0}, e | e, {st = Some 0} -> e
  | {st = Some m}, {st = Some n} -> int (R.add m n)
  | _ -> pdyn (C.add (abstr e1) (abstr e2))
  let if_ eb et ee = match eb with
  | {st = Some b} -> if b then et () else ee ()
  | _ -> pdyn (C.if_ (abstr eb) (fun () -> abstr (et ()))
                                (fun () -> abstr (ee ())))
  let lam f = {st = Some f; dy = C.lam (fun x -> abstr (f (pdyn x)))}
  let app ef ea = match ef with {st = Some f} -> f ea
                  | _ -> pdyn (C.app (abstr ef) (abstr ea)) end
\end{code2}
\medskip
\end{floatrule}
\caption{Our partial evaluator (\ifshort \texttt{bool}, \texttt{mul},
  \texttt{leq} and \texttt{fix} \else \texttt{mul} and \texttt{leq} \fi
  are elided)}
\label{fig:pe}
\end{figure}

@@

Any partial evaluator must decide how much to unfold recursion
statically: unfolding too little can degrade the residual code, whereas
unfolding too much risks nontermination.  Our partial evaluator is no
exception, because our object language includes |fix|.  The code in
Figure~\ref{fig:pe} takes the na\"\i ve approach of ``going all the
way'', that is, whenever the 
argument is static, we unfold |fix| rather than residualize it.
In the accompanying source code is a conservative
alternative |P.fix| that unfolds recursion only once, then residualizes.
Many sophisticated approaches have been developed to decide how much to unfold
\citep{jones-partial}, but this issue is orthogonal to our presentation.
\ifshort\else
A separate concern in our treatment of |fix| is possible code bloat in
the residual program, which calls for let-insertion
\citep{SwadiTahaKiselyovPasalic2006}.
\fi

@@

\section{A tagless partial evaluator}\label{PE}
Surprisingly, we can write a partial evaluator using the idea above,
namely to build object terms using ordinary functions rather than data
constructors.  We present this partial evaluator in a sequence of three
attempts. It uses no universal type and no tags
for object types.  We then discuss residualization and binding-time
analysis.  Our partial evaluator is a modular extension of the evaluator
in~\S\ref{S:interpreter-RL} and the compiler in~\S\ref{S:compiler}, in
that it uses the former to reduce static terms and the latter to build
dynamic terms.

@@

Roughly, a partial evaluator interprets each object term to yield either
a static (present-stage) term (using the evaluator~|R|) or
a dynamic (future-stage) term (using the compiler~|C|).  To
distinguish between static and dynamic terms, we might try to define
|repr| in the partial evaluator (module |P0|) as
\begin{code}
type ('c,'dv) repr = S0 of ('c,'dv) R.repr | D0 of ('c,'dv) C.repr
\end{code}
(the digit zero in the names of the phase tags |S0| and |D0| is the
version suffix).
To extract a dynamic term from this type, we create the function
\begin{code}
let abstrI0 (e:('c,int) repr) : ('c,int) C.repr =
  match e with
  | S0 e -> C.int e
  | D0 e -> e
\end{code}
and the similar function |abstrB0| for dynamic boolean terms. Here, 
|C.int| is used to convert from the static term |('c,int) R.repr|, 
which is just |int|, to the dynamic term. We can now define the following
components of module |P0| required by the |Semantics| signature:
\begin{code}
let int  (x:int)  = S0 (R.int x)
let bool (x:bool) = S0 (R.bool x)
let add e1 e2 = 
  match (e1,e2) with
  | (S0 e1, S0 e2) -> S0 (R.add e1 e2)
  | (e1, e2)       -> D0 (C.add (abstrI0 e1) (abstrI0 e2))
\end{code}
Integer and boolean literals are immediate, present-stage
values. Addition yields a static term (using~|R.add|) if and only 
if both operands are static; otherwise we extract the dynamic terms 
from the operands and add them using~|C.add|.

Whereas |mul| and |leq| are as easy to define as |add|, we encounter
a problem with |if_|.  Suppose that the first argument to |if_| 
is a dynamic term
(of type |('c,bool) C.repr|), the second a static term 
(of type |('c,'a) R.repr|), and the third a
dynamic term (of type |('c,'a) C.repr|). We then need to convert
the static term to dynamic, but there is no polymorphic ``lift''
function, of type |'a -> ('c,'a) C.repr|, to send a value to the future stage
\citep{xi-guarded,WalidPOPL03}.
\begin{comment}
(By the way, if we
were to add polymorphic \texttt{lift} to the type class
\texttt{Symantics repr}, then \texttt{repr} would become an instance of
\texttt{Applicative} and thus \texttt{Functor}:\texttt{ fmap
f = app (lift f)~}.)
\end{comment}

@@

We start building the partial evaluator anew (module |P1|) and
switch to the data type
\begin{code}
type ('c,'dv) repr = P1 of ('c,'dv) R.repr option * ('c,'dv) C.repr
\end{code}
so that a partially evaluated term always contains a dynamic
component and sometimes contains a static component.  By 
distributivity, the two
alternative constructors of an |option| value, |Some| and |None|,
tag each partially evaluated term with a phase: either present or
future.  This tag is not an object type tag: all pattern matching below
is exhaustive. Because the future-stage component is always available, we
can now define the polymorphic function
|let abstr1 (P1 (_,dyn)) = dyn| of type
\texttt{('c,'dv) repr -> ('c,'dv) C.repr}
to extract it without requiring polymorphic lift into~|C|.  We then try
to define the interpreter |P1|---and get as far as the first-order
constructs of our object language, including |if_|.
\begin{code}
module P1 : Symantics = struct
  let int (x:int) = P1 (Some (R.int x), C.int x)
  let add e1 e2 = match (e1,e2) with
    | (P1 (Some n1,_),P1 (Some n2,_)) -> int (R.add n1 n2)
    | _ -> P1 (None,(C.add (abstr1 e1) (abstr1 e2)))
  let if_ = function
    | P1 (Some s,_) -> fun et ee -> if s then et () else ee ()
    | eb -> fun et ee -> P1 (None, C.if_ (abstr1 eb) 
                                   (fun () -> abstr1 (et ()))
                                   (fun () -> abstr1 (ee ())))
\end{code}
However, we stumble on functions.  According to our
definition of~|P1|, a partially evaluated object function, such as the
identity $\fun{x}x$ embedded in OCaml as |lam (fun x -> x)|\texttt{ :
}|('c,'a->'a) P1.repr|, consists of a dynamic part 
(type |('c,'a->'a) C.repr|) and
maybe a static part (type |('c,'a->'a) R.repr|).  The dynamic part is useful
when this function is passed to another function that is only
dynamically known, as in $\fun{k}k(\fun{x}x)$.  The static part is
useful when this function is applied to a static argument, as in
$(\fun{x}x)\True$.  Neither part, however, lets us \emph{partially}
evaluate the function, that is, compute as much as possible statically
when it is applied to a mix of static and dynamic inputs.  For example,
the partial evaluator should turn $\fun{n}(\fun{x}x)n$ into $\fun{n}n$
by substituting $n$ for~$x$ in the body of $\fun{x}x$ even though $n$ is
not statically known.  The same static function, applied to
different static arguments, can give both static and dynamic results: we
want to simplify $(\fun{y}x\times y)0$ to~$0$ but $(\fun{y}x\times y)1$
to~$x$.

To enable these simplifications, we delay binding-time analysis
for a static function until it is applied, that is, until |lam f|
appears as the argument of |app|.  To do so, we have to incorporate |f|
as it is into the |P1.repr| data structure: the representation
for a function type |'a->'b| should be one of
\begin{code}
S1 of ('c,'a) repr -> ('c,'b) repr | E1 of ('c,'a->'b) C.repr
P1 of (('c,'a) repr -> ('c,'b) repr) option * ('c,'a->'b) C.repr
\end{code}
unlike |P1.repr| of |int| or |bool|.
That is, we need a nonparametric data type, something akin to
type-indexed functions and type-indexed types, which
\citet{oliveira-typecase} dub the \emph{typecase} design pattern.
Thus, typed partial evaluation, like typed CPS transformation
(see \S\ref{S:CPS}),
inductively defines a map from source types to target types that
performs case distinction on the source type. In Haskell, typecase
can be equivalently implemented either with GADTs or with
type-class functional dependencies
\citep{oliveira-typecase}. The accompanying code shows both
approaches (|Incope.hs| and |incope1.hs|), 
neither portable to OCaml. In addition,
the problem of nonexhaustive pattern\hyp matching reappears in
the GADT approach because GHC 6.6.1 cannot see that a particular
type of a GADT value precludes certain constructors.
Thus GADTs fail to
make it \emph{syntactically} apparent that pattern matching is exhaustive.

%% \subsection{Eliminating tags from typecase}
%% \label{S:PE-GADT}

%% Two common ways to provide typecase in Haskell are
%% GADTs and type-class functional dependencies
%% \citep{oliveira-typecase}.  These
%% methods are equivalent, and here we use GADTs; |incope1.hs|
%% in the accompanying source code shows the latter.
%% We introduce a GADT with four data constructors.
%% \begin{code}
%% data P t where
%%   VI :: Int  -> P Int
%%   VB :: Bool -> P Bool
%%   VF :: (P a -> P b) -> P (a -> b)
%%   E  :: C t -> P t
%% \end{code}
%% The constructors |VI|, |VB|, and |VF| build static terms (like |S0|
%% in~\S\ref{S:PE-lift}), and |E| builds dynamic terms (like |D0|).  However,
%% the type |P t| is no longer parametric in~|t|: the constructor |VF| takes an
%% operand of type |P a -> P b| rather than |a -> b|.  We define a function
%% like |abstr1| above to extract a future-stage computation from a 
%% value of type |P t|.
%% \begin{code}
%% abstr :: P t -> C t
%% abstr (VI i) = int i
%% abstr (VB b) = bool b
%% abstr (VF f) = lam (abstr . f . E)
%% abstr (E x)  = x
%% \end{code}
%% The cases of this function |abstr| are type-indexed.  In particular, the |VF f|
%% case uses the method |lam| of the |C| interpreter to compile~|f|.

%% We may now make |P| an instance of
%% |Symantics| and implement the partial evaluator as follows. We elide
%% |mul|, |leq|, |if_|, and |fix|.
%% \begin{code}
%% instance Symantics P where
%%   int x               = VI x
%%   bool b              = VB b
%%   add (VI n1) (VI n2) = VI (n1 + n2)
%%   add e1 e2           = E (add (abstr e1) (abstr e2))
%%   lam                 = VF
%%   app (VF f) ea       = f ea
%%   app (E f)  ea       = E (app f (abstr ea))
%% \end{code}
%% The implementations of |int|, |bool|, and |add| are like 
%% in~\S\ref{S:PE-problem}. The interpretation of |lam f| is |VF f|, 
%% which just wraps the HOAS function |f|. 
%% We can always compile |f| to a code value,
%% but we delay it to apply |f| to concrete arguments. The interpretation of
%% |app ef ea| checks to see if |ef| is such a delayed
%% HOAS function |VF f|. If it is, we apply |f| to the
%% concrete argument |ea|, giving us a chance to perform static
%% computations (see example |testpowfix7| in~\S\ref{S:PE-solution}). If |ef| is a
%% dynamic value |E f|, we residualize.

%% This solution using GADTs works but is not quite satisfactory. First, it
%% cannot be ported to MetaOCaml, as GADTs are unavailable there.  Second,
%% the problem of nonexhaustive pattern\hyp matching reappears in
%% |app| above: the type |P t| has four constructors, of which the pattern in
%% |app| matches only |VF| and~|E|. One may say that the
%% constructors |VI| and |VB| obviously cannot occur because they do not
%% construct values of type |P (a -> b)| as required by the type of |app|.
%% Indeed, the metalanguage implementation could reason thus:
%% if we use inductive families (as in Coq) or logical
%%   frameworks with canonical forms (as in Twelf with its coverage checker),
%%   we can prove the pattern matching to be exhaustive.
%% Then again, the metalanguage implementation may not reason thus:
%% GHC cannot and issues warnings.
%% Although this point may seem minor, it is the heart of
%% the tagging problem and the purpose of tag elimination. A typed tagged
%% interpreter contains many pattern\hyp matching forms that look partial
%% but never fail in reality. The
%% goal is to make this exhaustiveness \emph{syntactically} apparent.

@@

Let us re-examine the problem in~\S\ref{S:PE-problem}. What we
would ideally like is to write
\begin{code}
type ('c,'dv) repr = P1 of (repr_pe ('c,'dv)) R.repr option * ('c,'dv) C.repr
\end{code}
where |repr_pe| is the type function defined
% inductively because P below depends on repr_pe
by
\begin{code}
repr_pe ('c,int)    = ('c,int)
repr_pe ('c,bool)   = ('c,bool)
repr_pe ('c,'a->'b) = ('c,'a) repr -> ('c,'b) repr
\end{code}
Although we can use type classes to define this type function
in Haskell, that is not portable to MetaOCaml. However,
these three typecase alternatives are already present in existing
methods of |Symantics|.
A simple and portable solution thus emerges: we bake |repr_pe| 
into the signature |Symantics|. 
We recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that 
the |repr| type constructor took two arguments |'c| and~|'dv|.
We add an argument |'sv| for the result of applying |repr_pe| to~|'dv|.
Figure~\ref{fig:ocaml} shows the new signature.
\begin{figure}
\begin{floatrule}
\begin{code2}
module type Symantics = sig
  type ('c,'sv,'dv) repr

@@

  let lam f = {st = Some f; dy = C.lam (fun x -> abstr (f (pdyn x)))}
  let app ef ea = match ef with
                  | {st = Some f} -> f ea
                  | _ -> pdyn (C.app (abstr ef) (abstr ea))
  let fix f = let fdyn = C.fix (fun x -> abstr (f (pdyn x)))
              in let rec self = function
                                | {st = Some @!_} as e -> app (f (lam self)) e
                                | e -> pdyn (C.app fdyn (abstr e))
                 in {st = Some self; dy = fdyn}
end
\end{code2}
\end{floatrule}
\caption{Our partial evaluator (\texttt{mul} and \texttt{leq}
  are elided)}
\label{fig:pe}
\end{figure}

@@
